---
title: "Analysis of Spatial & Temporal Data - Final Project"
subtitle: <span style="color:dodgerblue">Air Quality Dataset - Indian Cities (2022-2025)</span>
author: 
  - Nguyen The Khai – 23BI14205
  - Chau Tuan Minh  - 23BI14299
  - Tran Gia Khanh  - 23BI14218
  - Ngo Xuan Kien   - 23BI14239
  - Ngo Minh Phuoc  - 23BI14361
  - Nguyen Thanh Vinh - 23BI14453
  - Tran Ngoc Hung    - 23BI14181
affiliation: "Data Science Major"
date: "`r Sys.Date()`"
output: 
  rmdformats::readthedown:
    highlight: espresso
    code_download: FALSE
    code_folding: show
    number_sections: TRUE
editor_options: 
  markdown: 
    wrap: sentence
---

# Environment setting

```{r}
library(tidyverse)
library(grid)
library(gridExtra)
library(plotly)
library(lubridate)
library(sf)
library(ggplot2)

library(RKaggle)
library(dplyr)
library(leaflet)
```

# Introduction

# Dataset Description

-   A comprehensive hourly dataset tracking atmospheric conditions and air quality across 29 major Indian cities (28 states + Delhi) from 2022 to 2025.
    This dataset provides detailed insights into India's environmental patterns, pollution trends, and meteorological conditions.

-   Geographic Coverage

    -   Cities Covered: 29 major cities representing all 28 Indian states + Delhi
    -   Time Period: 2022 to 2025
    -   Records: 842,160 hourly observations
    -   Features: 63 parameters covering weather, pollution, and temporal data

-   Dataset Specifications

    -   Total Records: 842,160
    -   Features: 63 columns
    -   Temporal Resolution: Hourly data
    -   Period: 2022-2025 (4 years)

-   Indian Cities Included

(Representing all 28 states + Union Territory of Delhi)

```         
- Delhi
- Mumbai (Maharashtra)
- Chennai (Tamil Nadu)
- Kolkata (West Bengal)
- Bengaluru (Karnataka)
- Hyderabad (Telangana)
- Ahmedabad (Gujarat)
- Pune (Maharashtra)
- Jaipur (Rajasthan)
- Lucknow (Uttar Pradesh)
- And other major state capitals…
```

-   Acknowledgments

Data compiled from various monitoring networks across India including:

```         
- Central Pollution Control Board (CPCB)
- Indian Meteorological Department (IMD)
- System of Air Quality and Weather Forecasting And Research (SAFAR)
```

## Data Source / Structure

-   [Air Quality Dataset: Indian Cities (2022-2025)](https://www.kaggle.com/datasets/bhautikvekariya21/air-quality-dataset-indian-cities-2022-2025?utm_source=chatgpt.com&select=INDIA_AQI_COMPLETE_20251126.csv)

```{r}
dat = get_dataset("bhautikvekariya21/air-quality-dataset-indian-cities-2022-2025")
dat = dat[[2]]
```

```{r}
glimpse(dat)
class(dat)
names(dat)
```

-   Data types:

    -   Spatial: City, State, Latitude, Longitude

    -   Temporal: Datetime, Year, Month, Hour, Season

    -   Pollution: PM2_5_ugm3, PM10_ugm3, AQI

    -   Meteorology: temperature, humidity, wind, rain

# Data Preprocessing

## Variable selection

```{r}
dat_clean <- dat %>%
  select(
    City, State, Latitude, Longitude,
    Datetime, Year, Month, Day, Hour,
    Season, Time_of_Day,
    Temp_2m_C, Humidity_Percent, Wind_Speed_10m_kmh,
    Rain_mm, Is_Raining,
    PM2_5_ugm3, PM10_ugm3, US_AQI,
    Is_Weekend, Crop_Burning_Season, Temp_Inversion,
  )

head(dat_clean)
```

## Missing values handling

```{r}
dat_clean <- dat_clean %>%
  filter(!is.na(US_AQI))

colSums(is.na(dat_clean))
```

```         
- Rows with missing US AQI values (145 observations, <0.02% of the dataset) were removed to avoid introducing artificial values into the time series.
```

# EDA

## Temporal Analysis

### Hourly pattern

This line plot shows the average PM2.5 concentration for each hour (0–23) across all observations

```{r}
hourly_pm <- dat_clean %>%
  group_by(Hour) %>%
  summarise(
    mean_pm25 = mean(PM2_5_ugm3, na.rm = TRUE)
  )

ggplot(hourly_pm, aes(x = Hour, y = mean_pm25)) +
  geom_line() +
  labs(
    title = "Average PM2.5 by Hour of Day",
    x = "Hour",
    y = "PM2.5 (ug/m^3)"
  )
```

```         
- PM2.5 is higher during late evening and night (19:00–22:00)and the lowest concentration occurs in early afternoon (13:00–15:00)
- Nighttime accumulation due to low wind speed and temperature inversion
- Afternoon reduction likely caused by strong sunlight and atmospheric mixing
- The pattern reflects a diurnal pollution cycle
```

### Monthly pattern

This plot shows the average PM2.5 concentration by month (1–12)

```{r}
monthly_pm <- dat_clean %>%
  group_by(Month) %>%
  summarise(
    mean_pm25 = mean(PM2_5_ugm3, na.rm = TRUE)
  )

ggplot(monthly_pm, aes(x = Month, y = mean_pm25)) +
  geom_line() +
  labs(
    title = "Monthly Average PM2.5",
    x = "Month",
    y = "PM2.5 (ug/m^3)"
  )
```

```         
- PM2.5 is highest in January and December
- Pollution gradually decreases from January to July
- The lowest level appears around July (monsoon period)
- Pollution increases again from October onward
- This indicates strong seasonal effects, likely influenced by: Winter temperature inversion, Post-monsoon stagnation or Possible crop burning season
```

### Seasonal Pattern

Boxplot showing PM2.5 distribution across four seasons:

```         
- Monsoon
- Post-Monsoon
- Summer
- Winter
```

```{r}
ggplot(dat_clean, aes(x = Season, y = PM2_5_ugm3)) +
  geom_boxplot() +
  labs(
    title = "PM2.5 Distribution by Season",
    x = "Season",
    y = "PM2.5 (ug/m^3)"
  )
```

```         
- Winter shows higher median and wider spread, meaning more extreme pollution events.
- Monsoon has lowest median, indicating rainfall reduces pollution.
- Summer shows large outliers, meaning occasional extreme events.
- High number of outliers suggests pollution spikes in certain conditions.
```

## Spatial Analysis

### City-level Average PM2.5

```{r}
city_pm <- dat_clean %>%
  group_by(City, State, Latitude, Longitude) %>%
  summarise(
    mean_pm25 = mean(PM2_5_ugm3, na.rm = TRUE),
    .groups = "drop"
  )
```

### Spatial Scatter

Scatter plot of cities by:

```         
- Longitude (x-axis)
- Latitude (y-axis)
- Color = Average PM2.5 level
```

```{r}
ggplot(city_pm, aes(x = Longitude, y = Latitude, color = mean_pm25)) +
  geom_point() +
  labs(
    title = "Spatial Distribution of Average PM2.5",
    x = "Longitude",
    y = "Latitude",
    color = "PM2.5"
  )
```

```         
- Northern and central regions appear to have higher PM2.5 values.
- Some clusters indicate regional pollution hotspots.
- Spatial differences shows pollution is not evenly distributed geographically.
```

# Weather & Pollution Relationship

## Correlation Analysis

### Temperature & PM2.5

Scatter plot showing relationship between:Temperature (oC) & PM2.5 concentration

```{r}
ggplot(dat_clean, aes(x = Temp_2m_C, y = PM2_5_ugm3)) +
  geom_point(alpha = 0.3) +
  labs(
    title = "PM2.5 vs Temperature",
    x = "Temperature (oC)",
    y = "PM2.5 (ug/m^3)"
  )
```

```         
- No strong linear relationship is clearly visible
- Higher PM2.5 values appear more frequently at moderate temperatures (10–25oC)
- Extreme pollution events seem less common at very high temperatures
- Temperature alone might not explain pollution, consider interaction with other variables (humidity, wind, season)
```

### Humidity vs PM2.5

Scatter plot showing relationship between: Humidity (%) & PM2.5 concentration

```{r}
ggplot(dat_clean, aes(x = Humidity_Percent, y = PM2_5_ugm3)) +
  geom_point(alpha = 0.3) +
  labs(
    title = "PM2.5 vs Humidity",
    x = "Humidity (%)",
    y = "PM2.5 (ug/m^3)"
  )
```

```         
- PM2.5 appears to increase as humidity increases
- Higher pollution concentrations are observed at high humidity levels
```

This may indicate:

```         
- Moist air traps pollutants
- Secondary particle formation under humid conditions
- However, variation is still large, leads to relationship is not perfectly linear
```

### Rain effect

Boxplot comparing PM2.5 on: No rain (0), Rain(1)

```{r}
dat_clean %>%
  group_by(Is_Raining) %>%
  summarise(
    mean_pm25 = mean(PM2_5_ugm3, na.rm = TRUE)
  )
ggplot(dat_clean, aes(x = factor(Is_Raining), y = PM2_5_ugm3)) +
  geom_boxplot() +
  labs(
    title = "Effect of Rain on PM2.5",
    x = "Is Raining (0 = No, 1 = Yes)",
    y = "PM2.5 (ug/m^3)"
  )
```

```         
- Rainfall has a cleansing effect (wet deposition removes particles)
- Median PM2.5 is clearly lower on rainy days
- Distribution is tighter when raining
- Non-rainy days show more extreme pollution outliers
```

## Conditional Analysis

# AQI Category Analysis

```{r}
ggplot(dat_clean, aes(x = US_AQI)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of US AQI",
    x = "US AQI",
    y = "Frequency"
  )
```

# City comparison

```{r}
dat_clean %>%
  filter(City %in% c("Delhi", "Mumbai")) %>%
  group_by(City, Month) %>%
  summarise(
    mean_pm25 = mean(PM2_5_ugm3, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = Month, y = mean_pm25, color = City)) +
  geom_line() +
  labs(
    title = "Monthly PM2.5 Comparison: Delhi vs Mumbai",
    x = "Month",
    y = "PM2.5 (ug/m^3)"
  )
```

# Temporal Decomposition

```{r}
dat_clean <- dat %>%
  select(
    City, State, Latitude, Longitude,
    Datetime, Year, Month, Day, Hour,
    Season, Time_of_Day,
    Is_Weekend,
    Crop_Burning_Season,
    Festival_Period,
    Temp_Inversion,
    Temp_2m_C, Humidity_Percent, Wind_Speed_10m_kmh,
    Rain_mm, Is_Raining,
    PM2_5_ugm3, PM10_ugm3, US_AQI
  )

table(dat_clean$Temp_Inversion)
unique(dat_clean$Temp_Inversion)
table(dat_clean$Crop_Burning_Season)
table(dat_clean$Is_Weekend)
```

## Trend Over Years

```{r}
yearly_pm <- dat_clean %>%
  group_by(Year) %>%
  summarise(mean_pm25 = mean(PM2_5_ugm3))

ggplot(yearly_pm, aes(x = Year, y = mean_pm25)) +
  geom_line() +
  labs(title = "Yearly PM2.5 Trend")
```

## Weekday & Weekend Analysis

```{r}
dat_clean %>%
  group_by(Is_Weekend) %>%
  summarise(mean_pm25 = mean(PM2_5_ugm3))

ggplot(dat_clean, aes(x = factor(Is_Weekend), y = PM2_5_ugm3)) +
  geom_boxplot() +
  labs(
    title = "Weekend vs Weekday PM2.5",
    x = "0 = Weekday, 1 = Weekend",
    y = "PM2.5"
  )
```

## Diurnal Cycle by Season

```{r}
dat_clean %>%
  group_by(Season, Hour) %>%
  summarise(mean_pm25 = mean(PM2_5_ugm3)) %>%
  ggplot(aes(x = Hour, y = mean_pm25, color = Season)) +
  geom_line()
```

# Extreme Events

## Set High Pollution Threshold

```{r}
high_pollution <- dat_clean %>%
  filter(PM2_5_ugm3 > 100)
```

## Top Most Polluted Cities

```{r}
city_rank <- dat_clean %>%
  group_by(City, State) %>%
  summarise(
    mean_pm25 = mean(PM2_5_ugm3),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_pm25)) %>%
  mutate(Rank = row_number())

top10 = city_rank %>%
  slice(1:10)

top10
```

```{r}
ggplot(top10, aes(x = reorder(City, mean_pm25), y = mean_pm25)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 Most Polluted Cities (Average PM2.5)",
    x = "City",
    y = "Mean PM2.5"
  )
```

# Typical Event Exploring

## Crop Burning Effect

```{r}
table(dat_clean$Crop_Burning_Season)

dat_clean %>%
  group_by(Crop_Burning_Season) %>%
  summarise(mean_pm25 = mean(PM2_5_ugm3))
```

# Correlation Matrix

```{r}
num_vars <- dat_clean %>%
  select(PM2_5_ugm3, Temp_2m_C, Humidity_Percent, Wind_Speed_10m_kmh, Rain_mm)

cor(num_vars, use = "complete.obs")
```

# Spatial Clustering

```{r}
ggplot(city_pm, aes(Longitude, Latitude, size = mean_pm25)) +
  geom_point()
```

# Leaflet Interactive Map

```{r}
leaflet_data <- dat_clean %>%
  group_by(City, State, Latitude, Longitude) %>%
  summarise(
    mean_pm25 = mean(PM2_5_ugm3),
    mean_aqi = mean(US_AQI, na.rm = TRUE),
    .groups = "drop"
  )

pal <- colorNumeric(
  palette = "YlOrRd",
  domain = leaflet_data$mean_pm25
)

leaflet(leaflet_data) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~Longitude,
    lat = ~Latitude,
    radius = 6,
    fillColor = ~pal(mean_pm25),
    fillOpacity = 0.8,
    stroke = FALSE,
    popup = ~paste(
      "<b>City:</b>", City, "<br>",
      "<b>Mean PM2.5:</b>", round(mean_pm25, 2), "<br>",
      "<b>Mean AQI:</b>", round(mean_aqi, 2)
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~mean_pm25,
    title = "Mean PM2.5",
    position = "bottomright"
  )
```

# Discussion & Conclusion
