---
title: "Analysis of Spatial & Temporal Data - Final Project"
subtitle: <span style="color:dodgerblue">Air Quality Dataset - Indian Cities (2022-2025)</span>
author: 
  - Nguyen The Khai – 23BI14205
  - Chau Tuan Minh  - 23BI14299
  - Tran Gia Khanh  - 23BI14218
  - Ngo Xuan Kien   - 23BI14239
  - Ngo Minh Phuoc  - 23BI14361
  - Nguyen Thanh Vinh - 23BI14453
  - Tran Ngoc Hung    - 23BI14181
affiliation: "Data Science Major"
date: "`r Sys.Date()`"
output: 
  rmdformats::readthedown:
    highlight: espresso
    code_download: FALSE
    code_folding: show
    number_sections: TRUE
editor_options: 
  markdown: 
    wrap: sentence
---

# Environment setting

```{r}
library(tidyverse)
library(grid)
library(gridExtra)
library(plotly)
library(lubridate)
library(sf)
library(ggplot2)

library(RKaggle)
```

# Introduction

# Dataset Description
- A comprehensive hourly dataset tracking atmospheric conditions and air quality across 29 major Indian cities (28 states + Delhi) from 2022 to 2025. This dataset provides detailed insights into India's environmental patterns, pollution trends, and meteorological conditions.

- Geographic Coverage

    - Cities Covered: 29 major cities representing all 28 Indian states + Delhi
    - Time Period: 2022 to 2025
    - Records: 842,160 hourly observations
    - Features: 63 parameters covering weather, pollution, and temporal data

- Dataset Specifications

    - Total Records: 842,160
    - Features: 63 columns
    - Temporal Resolution: Hourly data
    - Period: 2022-2025 (4 years)

- Indian Cities Included

(Representing all 28 states + Union Territory of Delhi)

    - Delhi
    - Mumbai (Maharashtra)
    - Chennai (Tamil Nadu)
    - Kolkata (West Bengal)
    - Bengaluru (Karnataka)
    - Hyderabad (Telangana)
    - Ahmedabad (Gujarat)
    - Pune (Maharashtra)
    - Jaipur (Rajasthan)
    - Lucknow (Uttar Pradesh)
    - And other major state capitals…

- Acknowledgments

Data compiled from various monitoring networks across India including:

    - Central Pollution Control Board (CPCB)
    - Indian Meteorological Department (IMD)
    - System of Air Quality and Weather Forecasting And Research (SAFAR)

## Data Source / Structure
- [Air Quality Dataset: Indian Cities (2022-2025)](https://www.kaggle.com/datasets/bhautikvekariya21/air-quality-dataset-indian-cities-2022-2025?utm_source=chatgpt.com&select=INDIA_AQI_COMPLETE_20251126.csv)

```{r}
dat = get_dataset("bhautikvekariya21/air-quality-dataset-indian-cities-2022-2025")
dat = dat[[2]]
```

```{r}
glimpse(dat)
class(dat)
names(dat)
```
- Data types:

    - Spatial: City, State, Latitude, Longitude

    - Temporal: Datetime, Year, Month, Hour, Season

    - Pollution: PM2_5_ugm3, PM10_ugm3, AQI

    - Meteorology: temperature, humidity, wind, rain
# Data Preprocessing

```{r}
dat_clean <- dat %>%
  select(
    City, State, Latitude, Longitude,
    Datetime, Year, Month, Day, Hour,
    Season, Time_of_Day,
    Temp_2m_C, Humidity_Percent, Wind_Speed_10m_kmh,
    Rain_mm, Is_Raining,
    PM2_5_ugm3, PM10_ugm3, US_AQI
  )

head(dat_clean)
```

## Missing values handling

```{r}
dat_clean <- dat_clean %>%
  filter(!is.na(US_AQI))

colSums(is.na(dat_clean))
```
    - Rows with missing US AQI values (145 observations, <0.02% of the dataset) were removed to avoid introducing artificial values into the time series.

## Variable selection

# EDA

## Temporal Analysis

### Hourly pattern
```{r}
hourly_pm <- dat_clean %>%
  group_by(Hour) %>%
  summarise(
    mean_pm25 = mean(PM2_5_ugm3, na.rm = TRUE)
  )

ggplot(hourly_pm, aes(x = Hour, y = mean_pm25)) +
  geom_line() +
  labs(
    title = "Average PM2.5 by Hour of Day",
    x = "Hour",
    y = "PM2.5 (ug/m^3)"
  )
```

### Monthly pattern 
```{r}
monthly_pm <- dat_clean %>%
  group_by(Month) %>%
  summarise(
    mean_pm25 = mean(PM2_5_ugm3, na.rm = TRUE)
  )

ggplot(monthly_pm, aes(x = Month, y = mean_pm25)) +
  geom_line() +
  labs(
    title = "Monthly Average PM2.5",
    x = "Month",
    y = "PM2.5 (ug/m^3)"
  )
```

### Seasonal Pattern 
```{r}
ggplot(dat_clean, aes(x = Season, y = PM2_5_ugm3)) +
  geom_boxplot() +
  labs(
    title = "PM2.5 Distribution by Season",
    x = "Season",
    y = "PM2.5 (ug/m^3)"
  )
```
## Spatial Analysis

### City-level Average PM2.5
```{r}
city_pm <- dat_clean %>%
  group_by(City, State, Latitude, Longitude) %>%
  summarise(
    mean_pm25 = mean(PM2_5_ugm3, na.rm = TRUE),
    .groups = "drop"
  )
```

### Spatial Scatter 

```{r}
ggplot(city_pm, aes(x = Longitude, y = Latitude, color = mean_pm25)) +
  geom_point() +
  labs(
    title = "Spatial Distribution of Average PM2.5",
    x = "Longitude",
    y = "Latitude",
    color = "PM2.5"
  )
```

# Weather & Pollution Relationship

## Correlation Analysis

### Temperature & PM2.5
```{r}
ggplot(dat_clean, aes(x = Temp_2m_C, y = PM2_5_ugm3)) +
  geom_point(alpha = 0.3) +
  labs(
    title = "PM2.5 vs Temperature",
    x = "Temperature (oC)",
    y = "PM2.5 (ug/m^3)"
  )
```

### Humidity vs PM2.5

```{r}
ggplot(dat_clean, aes(x = Humidity_Percent, y = PM2_5_ugm3)) +
  geom_point(alpha = 0.3) +
  labs(
    title = "PM2.5 vs Humidity",
    x = "Humidity (%)",
    y = "PM2.5 (ug/m^3)"
  )
```

### Rain effect 

```{r}
dat_clean %>%
  group_by(Is_Raining) %>%
  summarise(
    mean_pm25 = mean(PM2_5_ugm3, na.rm = TRUE)
  )
ggplot(dat_clean, aes(x = factor(Is_Raining), y = PM2_5_ugm3)) +
  geom_boxplot() +
  labs(
    title = "Effect of Rain on PM2.5",
    x = "Is Raining (0 = No, 1 = Yes)",
    y = "PM2.5 (ug/m^3)"
  )
```
## Conditional Analysis

# AQI Category Analysis

```{r}
ggplot(dat_clean, aes(x = US_AQI)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of US AQI",
    x = "US AQI",
    y = "Frequency"
  )
```
# City comparison

```{r}
dat_clean %>%
  filter(City %in% c("Delhi", "Mumbai")) %>%
  group_by(City, Month) %>%
  summarise(
    mean_pm25 = mean(PM2_5_ugm3, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = Month, y = mean_pm25, color = City)) +
  geom_line() +
  labs(
    title = "Monthly PM2.5 Comparison: Delhi vs Mumbai",
    x = "Month",
    y = "PM2.5 (ug/m^3)"
  )
```




